A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-08 summary.tex
added notes on future work

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

>>>>

\subsection{Generalization to locally polytropic disks}

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Higher order corrections}
%{\bf notes for self. maybe exclude in final. }

The simplified one-fluid formulation,
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\Delta\bm{v}|^2)$ when converting
the two-fluid equations to the one-fluid formulation. The full
one-fluid equations is significantly more complex than our 
model \cite[e.g.][]{laibe14}. However, it is possible to improve upon
the first-order formulation whilst retaining similarilty to  
standard hydrodynamics.

The full one-fluid equations of motion includes the acceleration term 
\begin{align}
%\left.\frac{D\bm{v}}{Dt}\right|_\mathrm{add} = 
\bm{a} =  
-\frac{1}{\rho}\nabla\cdot\left(\frac{\rhod\rhog}{\rho}\Delta\bm{v}\otimes\Delta\bm{v}\right) 
\equiv  \frac{1}{\rho}\nabla\cdot \bm{T},
\end{align}
where $\bm{T}$ is an effective stress tensor. 
In the terminal velocity approximation with $\Delta\bm{v} = \tstop
\nabla P/\rhog$, we have 
\begin{align}
\bm{T} = - \frac{\epsilon\tstop^2}{\rho}\nabla P \otimes \nabla P.
\end{align}
If we define 
\begin{align}
  \bm{g} \equiv \sqrt{4\pi G \epsilon \rho} \frac{\nabla P}{\rho}
  \tstop,\label{effective_g} 
\end{align}
then
\begin{align}
\bm{T} = - \frac{1}{4\pi G} \bm{g}\otimes\bm{g}, \label{effective_gstress}
\end{align}
%{\bf tensor virial theorems?}
which is just the anisotropic part of the  
gravitational stress tensor in self-gravitating hydrodynamics 
\citep{lynden-bell72}. However, unlike self-gravity, where $\bm{g}$ is
determined through the Poisson equation; here $\bm{g}$ is directly
related to $P$ (Eq. \ref{effective_g}), which itself obeys an 
evolutionary equation for isothermal gas (Eq. \ref{eff_energy}).

Nevertheless, Eq. \ref{effective_g}---\ref{effective_gstress}
indicates that pressure gradients provide torques for angular momentum
transport of the mixture as a whole. This is only possible for
non-axisymmetric pressure forces so that $T_{r\phi}\neq0$.

In a subsequent paper we will extend the one-fluid formulation by
including $\bm{T}$ in the equation of motion. 
In particular, we will explore whether this can surpress the suprious
modes discussed in Appendix \S\ref{spurious_epi}.

%stratified case require more boundary conditions
